Simulation of an Attic Firein a Wood Frame Residential Structure - Chicago, IL

### 2015-01-12 Studies/2012_Chicago/Report/firetech_chicago.tex
Chicago Fire: add firetech tex file

\documentclass[12pt,oneside]{book}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                                                 %
% The mathematical style of these documents follows                                               %
%                                                                                                 %
% A. Thompson and B.N. Taylor. The NIST Guide for the Use of the International System of Units.   %
%    NIST Special Publication 881, 2008.                                                          %
%                                                                                                 %
% http://www.nist.gov/pml/pubs/sp811/index.cfm                                                    %
%                                                                                                 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{../../../Bibliography/commoncommands}

% Rename chapter headings
\renewcommand{\chaptername}{Section}
\renewcommand{\bibname}{References}

% Math shortcuts
\renewcommand{\sb}[1]{_\mathrm{#1}}
\renewcommand{\C}{\mbox{C}}
\renewcommand{\H}{\mbox{H}}
\renewcommand{\O}{\mbox{O}}
\newcommand{\N}{\mbox{N}}

@@

% PDF metadata
\hypersetup{
  pdftitle={Simulation of an Attic Fire in a Wood Frame Residential Structure - Chicago, IL},
  pdfauthor={Weinschenk, Craig G, Overholt, Kristopher J., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2~\sfrac{1}{2}~story, single family structure in Chicago, IL, that resulted in the death of the captain of a fire department. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

\begin{document}
\pagenumbering{gobble}

\bibliographystyle{unsrt}
%\pagestyle{empty}

\begin{minipage}[t][9in][s]{6.25in}

\begin{flushright}
\fontsize{20}{24}\selectfont
\bf{NIST Technical Note 1838}
\end{flushright}

\headerB{
Simulation of an Attic Fire \\
in a Wood Frame Residential \\
Structure - Chicago, IL \\
}

\normalsize

\headerC{
{
\flushright{
Craig G. Weinschenk \\
Kristopher J. Overholt \\
Daniel Madrzykowski \\

\vspace*{2\baselineskip}

\begingroup
This publication is available free of charge from:
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.1838}{http://dx.doi.org/10.6028/NIST.TN.1838}
\endgroup
}

\vfill

\flushright{

\includegraphics[width=2.in]{../../../Bibliography/nistident_flright_vec} \\[.3in]
}
}
}

\end{minipage}

\newpage
\hspace{5in}
\newpage

\frontmatter

\begin{minipage}[t][9in][s]{6.25in}
\pagenumbering{gobble}

\begin{flushright}
\fontsize{20}{24}\selectfont
\bf{NIST Technical Note 1838}
\end{flushright}

\headerB{
Simulation of an Attic Fire \\
in a Wood Frame Residential \\
Structure - Chicago, IL \\
}

\headerC{
\flushright{
Craig G. Weinschenk \\
Kristopher J. Overholt \\
Daniel Madrzykowski \\
{\em Fire Research Division \\
Engineering Laboratory} \\

\vspace*{2\baselineskip}

\begingroup
This publication is available free of charge from:
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.1838}{http://dx.doi.org/10.6028/NIST.TN.1838} \\
\endgroup

\vspace*{2\baselineskip}

August 2014}}

\vfill

\flushright{\includegraphics[width=1in]{../../../Bibliography/doc} }

\titlesigs

\end{minipage}

\newpage

\begin{minipage}[t][9in][s]{6.25in}
\pagenumbering{gobble}

\flushright{Certain commercial entities, equipment, or materials may be identified in this \\
document in order to describe an experimental procedure or concept adequately. \\
Such identification is not intended to imply recommendation or endorsement by the \\
National Institute of Standards and Technology, nor is it intended to imply that the \\
entities, materials, or equipment are necessarily the best available for the purpose. \\
}

\vspace{3in}

\large
\flushright{\bf National Institute of Standards and Technology Technical Note 1838 \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~1838, \pageref{LastPage} pages (August 2014) \\
CODEN: NTNOEF}

\vspace{0.2in}

\begingroup
{\bf This publication is available free of charge from:}
\hypersetup{urlcolor=black}
\href{http://dx.doi.org/10.6028/NIST.TN.1838}{\bf http://dx.doi.org/10.6028/NIST.TN.1838} \\
\endgroup

\vfill

\hspace{1in}

\end{minipage}

\newpage

\frontmatter

%\pagestyle{plain}
\pagenumbering{roman}
\setcounter{page}{3}

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Contents}
\tableofcontents

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Tables}
\listoftables

\chapter{List of Acronyms}

\begin{tabbing}
\hspace{1.5in} \= \\
BC \> Battalion Chief \\
E \> Engine \\
FDS \> Fire Dynamics Simulator \\
HGL \> Hot Gas Layer \\
HRR \> Heat Release Rate \\
HRRPUA \> Heat Release Rate per Unit Area \\
IC \> Incident Command \\
LODD \> Line of Duty Death \\
NIOSH \> National Institute for Occupational Safety and Health \\
NIST \> National Institute of Standards and Technology \\
SFPE \> Society of Fire Protection Engineers \\
SQ \> Squad \\
TL \> Tower Ladder \\
\end{tabbing}

\chapter{List of Symbols}

\begin{tabbing}
\hspace{1.5in} \= \\
ft \> foot \\
m \> meter \\
min \> minute \\
Pa \> Pascal \\
psi \> pound per square inch \\
s \> second \\
W \> Watt \\
$k$ \> thermal conductivity \\
$\rho$ \> density \\
$c_{p}$ \> specific heat \\
\end{tabbing}

\mainmatter

\chapter*{\centering Abstract}
\pagenumbering{gobble}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2~\sfrac{1}{2}~story, single family structure in Chicago, IL, that resulted in the death of the captain of a fire department. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

\chapter{Introduction}
\pagenumbering{arabic}
\setcounter{page}{1}
Part of the function of the Fire Research Division at NIST is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of these studies has been to improve the safety and effectiveness of firefighters.

@@

\chapter{Summary of the Fire Incident}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines a rapid intervention team (RIT) response to be one truck company, one battalion chief, one advanced life support (ALS) ambulance, and one emergency medical services (EMS) field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

@@

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gases. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

In reality, the quantities associated with these input parameters are not fixed values; rather, a model input parameter can be thought of as a point estimate from a distribution of possible input parameters with some associated amount of uncertainty. Any change in an input parameter (such as the HRR) for a given scenario would result in a change in the output quantity (such as the hot gas layer (HGL) temperature). For example, according to the McCaffrey, Quintiere, and Harkleroad~\cite{SFPE:Walton} empirical correlation, the HGL temperature in a well-ventilated compartment fire is proportional to the HRR raised to the two-thirds power. Following this relationship, a 7.5~\% increase in the HRR would result in a 5~\% increase in the HGL temperature~\cite{NUREG_1824_Sup_1}. More detailed discussion on the propagation of parameter uncertainty in fire models is available in a validation study that was sponsored by the U.S. Nuclear Regulatory Commission~\cite{NUREG_1824_Sup_1}. The following subsections describe the inputs that were used to develop the simulation, including any approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

@@

\section{Fire}
\label{fire}

To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these sources of information, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board.

@@

\section{Materials}
\label{matl}
While the fires were represented as fuel sources with a constant area (Section~\ref{fire}), from a heat transfer perspective, it is important to define the material properties (density, thermal conductivity, and specific heat) to account for heat transfer and energy storage in the ceiling, walls, and floor. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings on the main two floors of the structure, and the material properties of wood~\cite{Incropera:1} were specified on the unfinished portions of the attic and rear porch. The following lines define these materials in the FDS input file:

@@

\section{Ventilation}
\label{Vents}
The simulations in this study account for changes in the ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). Two different ventilation scenarios were considered in this study. The first scenario involves a baseline simulation in which the ventilation times and ventilation areas represent our best understanding of the incident. The second scenario, or alternative simulation, does not include the firefighter opening the rear door to make entry to the structure and the subsequent exterior left wall fire penetration. The times of the ventilation changes are provided in Table~\ref{tab:vents}. Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents (opening size and open time) created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

@@

\section{Numerical Mesh}
\label{mesh}

@@

As discussed in Section~\ref{geom}, the structure has dimensions of 17.15~m (56~ft) by 6.15~m (20~ft) and a total height of 8.56~m (28~ft). To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The total volume of the computational domain was 24~m (79~ft) by 10~m (33~ft) by 10~m (33~ft). A grid resolution of 10~cm (3.9~in) was used, which required 2.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10~cm mesh. The domain was divided into 16 equally sized meshes (each containing 150,000 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 1.5~days. Figure~\ref{fig:mult_mesh} shows the entire structure within the computational domain that has been divided into multiple meshes. Table~\ref{tab:mod_param} shows a summary of the model input parameters for the simulations that were conducted as part of this study.

@@

\begin{table}
\centering
\captionof{table}{Relevant fire model input parameters}\label{tab:mod_param}
\begin{tabular}{lll}
\toprule[1.5pt]
Parameter                                                  & Description                                   & Discussion                            \\
\midrule
Simulation Time                                            &  5~min                                        & --                                    \\ [.25cm]
Grid Cell Size                                             &  10~cm                                        & Section~\ref{mesh}                    \\ [.25cm]
Ambient Temperature$^*$                                    &  20~$^{\circ}$C (68~$^{\circ}$F)              & --                                    \\ [.1cm]
\multirow{4}{*}{Reaction: Wood~\cite{SFPE:Tewarson}}       &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     & \multirow{4}{*}{Section~\ref{fire}}   \\
                                                           &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                       \\
                                                           &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                       \\
                                                           &  Heat of Combustion: 16,400~kJ/kg             &                                       \\ [.25cm]
\multirow{2}{*}{Peak HRR}                                  &  Attic Fire: 9~MW                             & \multirow{2}{*}{Section~\ref{fire}}   \\ 
                                                           &  Porch Fire: 3.3~MW                           &                                       \\ [.25cm]                     
\multirow{3}{*}{Material: Wood~\cite{Incropera:1} }        &  $k$: 0.12~\si{W/(m.K)})                      & \multirow{3}{*}{Section~\ref{matl}}   \\
                                                           &  $\rho$: 510~kg/m$^3$                         &                                       \\
                                                           &  $c_{p}$: 1.38~\si{kJ/(kg.K)}                 &                                       \\ [.25cm]
\multirow{3}{*}{Material: Gypsum Board~\cite{WAKILI2007}}  &  $k$: 0.28~\si{W/(m.K)}                       &  \multirow{3}{*}{Section~\ref{matl}}  \\ 
                                                           &  $\rho$: 810~kg/m$^3$                         &                                       \\ 
                                                           &  $c_{p}$: 1.0~\si{kJ/(kg.K)}                  &                                       \\
\bottomrule[1.25pt]
\end{tabular}\par
\footnotesize
$^{*}$Initial interior temperatures were assumed to be 20~$^{\circ}$C; exterior temperatures were 7~$^{\circ}$C~\cite{NIOSH:Bowyer}.
\normalsize
\end{table}

\chapter{Model Results}
\label{results}

@@

\chapter{Dimensioned Drawings}
\label{app_draw}

@@

\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof

\input{../../../Bibliography/commoncommands}
\usepackage{graphicx}
\renewcommand{\sb}[1]{_\mathrm{#1}}
\renewcommand{\C}{\mbox{C}}
\renewcommand{\H}{\mbox{H}}
\renewcommand{\O}{\mbox{O}}
\newcommand{\N}{\mbox{N}}

@@

% Insert the name of "your journal" with
\journalname{Fire Technology}
%
\begin{document}

\title{Simulation of an Attic Firein a Wood Frame Residential Structure - Chicago, IL}

%\titlerunning{Short form of title}        % if too long for running head

\author{Craig G. Weinschenk         \and
        Kristopher J. Overholt      \and
        Daniel Madrzykowski
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Craig Weinschenk \at
              National Instutue of Standards and Technology \\
              100 Bureau Drive \\
              Gaithersburg MD 20899-8661
              Tel.: 301-975-6899\\
              \email{craig.weinschenk@nist.gov}           %  \\
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2~\sfrac{1}{2}~story, single family structure in Chicago, IL, that resulted in the death of the captain of a fire department. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

\keywords{LODD \and FDS \and CFD}
\end{abstract}

\section{Introduction}
\label{intro}
Part of the function of the Fire Research Division at NIST is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of these studies has been to improve the safety and effectiveness of firefighters.

@@

\section{Summary of the Fire Incident}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines a rapid intervention team (RIT) response to be one truck company, one battalion chief, one advanced life support (ALS) ambulance, and one emergency medical services (EMS) field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

@@

\section{Model Description}
\label{model}
Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gases. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\subsection{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

@@

\subsection{Fire}
\label{fire}
To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these sources of information, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board.

@@

\subsection{Materials}
\label{matl}
While the fires were represented as fuel sources with a constant area (Section~\ref{fire}), from a heat transfer perspective, it is important to define the material properties (density, thermal conductivity, and specific heat) to account for heat transfer and energy storage in the ceiling, walls, and floor. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings on the main two floors of the structure, and the material properties of wood~\cite{Incropera:1} were specified on the unfinished portions of the attic and rear porch. The following lines define these materials in the FDS input file:

@@

\subsection{Ventilation}
\label{Vents}
The simulations in this study account for changes in the ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). Two different ventilation scenarios were considered in this study. The first scenario involves a baseline simulation in which the ventilation times and ventilation areas represent our best understanding of the incident. The second scenario, or alternative simulation, does not include the firefighter opening the rear door to make entry to the structure and the subsequent exterior left wall fire penetration. The times of the ventilation changes are provided in Table~\ref{tab:vents}. Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents (opening size and open time) created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

@@

\subsection{Numerical Mesh}
\label{mesh}

@@

As discussed in Section~\ref{geom}, the structure has dimensions of 17.15~m (56~ft) by 6.15~m (20~ft) and a total height of 8.56~m (28~ft). To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The total volume of the computational domain was 24~m (79~ft) by 10~m (33~ft) by 10~m (33~ft). A grid resolution of 10~cm (3.9~in) was used, which required 2.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10~cm mesh. The domain was divided into 16 equally sized meshes (each containing 150,000 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 1.5~days. Figure~\ref{fig:mult_mesh} shows the entire structure within the computational domain that has been divided into multiple meshes.

@@

\subsection{Summary of Model Input Parameters}
\label{sec:summary_of_model_inputs}

Table~\ref{tab:mod_param} shows a summary of the model input parameters for the simulation that was conducted as part of this study. In reality, the quantities associated with model input parameters are not fixed values; rather, a model input parameter can be thought of as a point estimate from a distribution of possible input parameters with some associated amount of uncertainty. Any change in an input parameter (such as the HRR) for a given scenario results in a change in the output quantity (such as the hot gas layer (HGL) temperature).

For example, according to the McCaffrey, Quintiere, and Harkleroad~\cite{SFPE:Walton} empirical correlation, the HGL temperature in a well-ventilated compartment fire is proportional to the HRR raised to the two-thirds power. Following this relationship, a 7.5~\% increase in the HRR would result in a 5~\% increase in the HGL temperature~\cite{NUREG_1824_Sup_1}. More detailed discussion on the propagation of parameter uncertainty in fire models is available in a validation study that was sponsored by the U.S. Nuclear Regulatory Commission~\cite{NUREG_1824_Sup_1}.

\begin{table}[h!]
\centering
\captionof{table}{Relevant fire model input parameters}\label{tab:mod_param}
\begin{tabular}{lll}
\toprule[1.5pt]
Parameter                                                  & Description                                   & Discussion                            \\
\midrule
Simulation Time                                            &  5~min                                        & --                                    \\ [.25cm]
Grid Cell Size                                             &  10~cm                                        & Section~\ref{mesh}                    \\ [.25cm]
Ambient Temperature$^*$                                    &  20~$^{\circ}$C (68~$^{\circ}$F)              & --                                    \\ [.1cm]
\multirow{4}{*}{Reaction: Wood~\cite{SFPE:Tewarson}}       &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     & \multirow{4}{*}{Section~\ref{fire}}   \\
                                                           &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                       \\
                                                           &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                       \\
                                                           &  Heat of Combustion: 16,400~kJ/kg             &                                       \\ [.25cm]
\multirow{2}{*}{Peak HRR}                                  &  Attic Fire: 9~MW                             & \multirow{2}{*}{Section~\ref{fire}}   \\ 
                                                           &  Porch Fire: 3.3~MW                           &                                       \\ [.25cm]                     
\multirow{3}{*}{Material: Wood~\cite{Incropera:1} }        &  $k$: 0.12~\si{W/(m.K)})                      & \multirow{3}{*}{Section~\ref{matl}}   \\
                                                           &  $\rho$: 510~kg/m$^3$                         &                                       \\
                                                           &  $c_{p}$: 1.38~\si{kJ/(kg.K)}                 &                                       \\ [.25cm]
\multirow{3}{*}{Material: Gypsum Board~\cite{WAKILI2007}}  &  $k$: 0.28~\si{W/(m.K)}                       &  \multirow{3}{*}{Section~\ref{matl}}  \\ 
                                                           &  $\rho$: 810~kg/m$^3$                         &                                       \\ 
                                                           &  $c_{p}$: 1.0~\si{kJ/(kg.K)}                  &                                       \\
\bottomrule[1.25pt]
\end{tabular}\par
\footnotesize
$^{*}$Initial interior temperatures were assumed to be 20~$^{\circ}$C; exterior temperatures were 7~$^{\circ}$C~\cite{NIOSH:Bowyer}.
\normalsize
\end{table}

\section{Model Results}
\label{results}

@@

\section{Dimensioned Drawings}
\label{app_draw}

@@

\end{document}
% end of file template.tex
